---
title: "GDP Components"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE }
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, echo = FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
```

Let's conduct an analysis to see how GDP and its components have changed
over time, and compare different countries and how much each component
contributes to that country's GDP. The main components of gross domestic
product are:

1.  C - personal consumption
2.  I - business investment (I)
3.  G - government spending
4.  Net exports (exports - imports).

The file we will work with is [GDP and its breakdown at constant 2010
prices in US Dollars](http://unstats.un.org/unsd/amaapi/api/file/6).
Let's have look at the dataset to see how it is structured and organised
using `read_excel` function:

```{r read_GDP_data}

UN_GDP_data  <-  read_excel(here::here("data", "Download-GDPconstant-USD-countries.xls"), 
                sheet="Download-GDPconstant-USD-countr", 
                skip=2) 
```

The data provided is is in wide format. Therefore, we need to make it
tidy by pivoting longer.

```{r reshape_GDP_data}

tidy_GDP_data <- UN_GDP_data %>% 
  pivot_longer(cols=4:51,
               names_to = "Year",
               values_to = "value")
glimpse(tidy_GDP_data)




```

I want to compare GDP components between 3 countries that I am
interested in. To do this, I am creating a vector with chosen countries.
Further, I mutate the value by dividing them by 1e9 so that the figures
are expressed in billions.

```{r }
country_list <- c("United States","India", "Germany")
tidy_GDP_data<-tidy_GDP_data %>% 
  filter(Country==country_list) %>% 
  mutate(value=value/10^9)
```

Further, I need to change indicator names to make my graph more
understandable. I am also changing variable Year for `integer` and
omitting missing values before I start plotting.

```{r gdp1_code}

tidy_GDP_for_graph1 <- tidy_GDP_data %>% 
  mutate(components_of_GDP=case_when(
    grepl("Exports",IndicatorName) ~ "Exports",
    grepl("Gross capital formation",IndicatorName) ~ "Gross capital formation",
    grepl("Government",IndicatorName) ~ "Government expenditure",
    grepl("government",IndicatorName) ~ "Government expenditure",
    grepl("Household",IndicatorName) ~ "Household expenditure",
    grepl("household",IndicatorName) ~ "Household expenditure",
    grepl("Imports",IndicatorName)~"Imports"
  )) %>% 
  mutate(Year = as.integer(Year)) %>% 
  na.omit(components_of_GDP)


```

Now we can start plotting our graph of evolution of the components of
GDP in Germany, India and United States.

```{r}
ggplot(tidy_GDP_for_graph1,
       aes(x=Year,y=value,color=components_of_GDP))+
  geom_line(aes(group=components_of_GDP))+
  facet_wrap(~Country)+
  labs(
    title = "Evolution of the components of GDP by country",
    subtitle = "In Constant 2010 USD",
    caption = "Source: United Nations' National Accounts Main Aggregates Database",
    color = "GDP Components",
    x = "Year",
    y = "Value in $M"
  )

```

As mentioned above, GDP is the sum of Household Expenditure (Consumption
*C*), Gross Capital Formation (business investment *I*), Government
Expenditure (G) and Net Exports (exports - imports).

Even though there is an indicator `Gross Domestic Product (GDP)` in your
dataframe, let's calculate it given its components discussed above.

First, we need to pivot out table wider to calculate GDP as sum of its
components, we will call it `wider_gdp`. Than, we will use `left_join`
function to connect the tables by country and year and to see the
percentage comparison.

```{r}
tidy_GDP_for_graph1 %>% 
  select(Country, components_of_GDP, Year, value) %>% 
  pivot_wider(names_from = components_of_GDP, values_from = value) %>% 
  janitor::clean_names() %>% 
  mutate(
    net_exports = exports - imports,
    gdp = household_expenditure + government_expenditure + gross_capital_formation + net_exports
    ) -> wider_gdp

wider_gdp %>% 
  select(country, year, gdp)

tidy_GDP_data_for_join = tidy_GDP_data%>% 
  filter(IndicatorName=="Gross Domestic Product (GDP)") %>% 
  mutate(Year=as.integer(Year)) %>% 
  select(Country,Year,value)
  
  
tidy_GDP_data_comparison=left_join(wider_gdp,tidy_GDP_data_for_join,by=c("country"="Country","year"="Year"))

tidy_GDP_data_comparison %>% 
  mutate(percentage_change=(value-gdp)/value) %>% 
  select(country,year,gdp,value,percentage_change)
```

> Above in the table we may see that there are some minor differences
> between what we calculated as GDP and the GDP figure included in the
> dataset.

Let's see what was the dynamic of GDP and its components in percentage
of GDP between the countries we chosen previously.

```{r}
wider_gdp %>% 
  mutate(
    percentage_GE = 100*government_expenditure/gdp,
    percentage_GCF = 100*gross_capital_formation/gdp,
    percentage_HE = 100*household_expenditure/gdp,
    percentage_NE = 100*net_exports/gdp
  ) %>% 
  select(country, year, percentage_GE, percentage_GCF, percentage_HE, percentage_NE) %>% 
  pivot_longer(cols = 3:6, names_to = "component_of_gdp", values_to = "value") %>% 
  ggplot(aes(x=year, y=value, color=component_of_gdp)) +
    geom_line() +
    facet_wrap(~country) +
    theme_minimal() +
    labs(
      title = "GDP and its breakdown at constant 2010 prices in US Dollars",
      x = NULL,
      y = "% of GDP",
      caption = "Source: United Nations",
      color = NULL
    ) +
    scale_color_hue(
      labels = c('Gross Capital Formation', 
                 'Government Expenditure', 
                 'Household Expenditure',
                 'Net Exports'
                 )
      )
```

> For all three countries household expenditure is the biggest
> contributor to the GDP, followed by gross capital formation,
> government expenditure, and finally Net Exports.
>
> The graph clearly shows that Germany is an exporting country, whereas
> the US seems to import more than it exports, and India seems to
> fluctuate close to zero.
>
> Between the three countries, India experiences the highest variability
> in GDP components, with Gross Capital Formation increasing greatly,
> and household expenditure decreasing greatly. In the United States,
> gross capital formation and government expenditure contributed roughly
> the same amount, with gross capital formation briefly increasing to
> more than government expenditure, before dipping back down to less
> than government expenditure.

Finally, let's plot a graph showing a dynamic of GDP and its components
between Poland, Spain, Singapore and China over given time period.

```{r}
country_list <- c("Poland","Spain", "Singapore", "China")

tidy_GDP_data <- UN_GDP_data %>% 
  pivot_longer(cols=4:51,
               names_to = "Year",
               values_to = "value")

tidy_GDP_data<-tidy_GDP_data %>% 
  filter(Country==country_list) %>% 
  mutate(value=value/10^9)

tidy_GDP_for_graph1 <- tidy_GDP_data %>% 
  mutate(components_of_GDP=case_when(
    grepl("Exports",IndicatorName) ~ "Exports",
    grepl("Gross capital formation",IndicatorName) ~ "Gross capital formation",
    grepl("Government",IndicatorName) ~ "Government expenditure",
    grepl("government",IndicatorName) ~ "Government expenditure",
    grepl("Household",IndicatorName) ~ "Household expenditure",
    grepl("household",IndicatorName) ~ "Household expenditure",
    grepl("Imports",IndicatorName)~"Imports"
  )) %>% 
  mutate(Year = as.integer(Year)) %>% 
  na.omit(components_of_GDP)

tidy_GDP_for_graph1 %>% 
  select(Country, components_of_GDP, Year, value) %>% 
  pivot_wider(names_from = components_of_GDP, values_from = value) %>% 
  janitor::clean_names() %>% 
  mutate(
    net_exports = exports - imports,
    gdp = household_expenditure + government_expenditure + gross_capital_formation + net_exports
    ) -> wider_gdp

wider_gdp %>% 
  mutate(
    percentage_GE = 100*government_expenditure/gdp,
    percentage_GCF = 100*gross_capital_formation/gdp,
    percentage_HE = 100*household_expenditure/gdp,
    percentage_NE = 100*net_exports/gdp
  ) %>% 
  select(country, year, percentage_GE, percentage_GCF, percentage_HE, percentage_NE) %>% 
  pivot_longer(cols = 3:6, names_to = "component_of_gdp", values_to = "value") %>% 
  ggplot(aes(x=year, y=value, color=component_of_gdp)) +
    geom_line() +
    facet_wrap(~country) +
    theme_minimal() +
    labs(
      title = "GDP and its breakdown at constant 2010 prices in US Dollars",
      x = NULL,
      y = "% of GDP",
      caption = "Source: United Nations",
      color = NULL
    ) +
    scale_fill_discrete(
      labels = c('Gross Capital Formation', 
                 'Government Expenditure', 
                 'Household Expenditure',
                 'Net Exports'
                 )
      )
```
